Classification and characterization of anthropogenic plant communities at the ecoregion level

### 2024-06-12 doc/manuscript.Rmd
submission-ready 🏙️

---
title: "Anthropogenic plant communities of the Iberian Atlantic ecoregion"
output: 
  bookdown::github_document2:
    reference_docx: template.docx
    number_sections: false
bibliography: refs.bib
csl: journal-of-vegetation-science.csl
link-citations: yes
always_allow_html: true
---

@@

@@

Anthropogenic vegetation is the set of plant communities that occur as a direct consequence of human activities, and is chiefly composed by the weed vegetation of arable fields and the ruderal vegetation of human settlements and their surroundings [@RN5441]. The anthropogenic communities of human-made habitats make up an important part of the supply of natural elements in urban and peri-urban habitats [@RN5445]. In densely-populated city landscapes, the plant communities that colonize urban vacant lots [@RN2300] are a useful biodiversity resource with high ecological and societal potential [@RN5446]. Anthropogenic vegetation cover can also impact regional-level extreme climatic events [@RN5440]. Furthermore, many synanthropic plants can accumulate heavy metals in their biomass [@RN5433], making them valuable candidates for nature-based solutions such as bioremediation and restoration of degraded landscapes [@RN5447].

The anthropogenic vegetation of temperate Europe is formed by an original pool of native species favored by humans, but this pool has been constantly enriched by the arrival of alien species: first archaeophytes starting in the Neolithic and peaking in the Bronze Age; and then neophytes from the Modern Age (after 1.500 A.D.) to the present day [@RN5439]. Synanthropic archaeophytes include many Mediterranean taxa that started their expansion into temperate areas during the Mid-Holocene warm period, but it is difficult to disentangle to what extent this expansion was favored by human activities and/or ongoing climatic changes [@RN5432; @RN5443], especially in southern European areas close to the transition between the Mediterranean and the temperate climates [@RN5444]. Meanwhile, synanthropic neophytes include many American, African and Asian taxa that arrived as a consequence of the global trade networks established by the Western European colonial powers [@RN5451]. In the last century, anthropogenic vegetation has shown a decrease in species richness and diversity [@RN5423], as rare synanthropic species have become rarer, and neophytes have increased their abundance at the expense of natives and archaeophytes [@RN5441].

Antropogenic vegetation is likely to be similar within ecological regions (ecoregions, hereafter), since they are assumed to encompass areas with a similar biogeographic history and recurrent local ecosystems [@RN5491]. However, ecoregional-scale studies on the diversity of antropogenic vegetation are scarce in the literature. The Iberian Atlantic ecoregion (a.k.a. Cantabrian Mixed Forests ecoregion; NW Iberian Peninsula) has a long history of human habitation dating back to the Cantabrian Upper Paleolithic [@RN5450]. This, together with its transitional position at the border between the temperate and Mediterranean climatic zones of Europe [@RN4700], suggests a long regional history for archaeophyte-rich anthropogenic plant communities, likely supporting a current high diversity of anthropogenic plant communities with both temperate and Mediterranean floristic elements [@RN5008]. Historical trade links with America and Asia, together with the warm and humid temperate climate, have made the region a hotspot for neophyte invasions [@RN5492; @RN5493]. Recent post-industrial land-use changes have left large areas with abandoned industrial sites (i.e. brownfields) whose management and restoration requires an understanding of anthropogenic communities able to colonize or restore them [@RN5448; @RN5449]. Although there is a long tradition of studying weed and ruderal communities in the ecoregion [@RN5364; @RN5361; @RN5359; @RN5360] there is a current need for an ecoregional-level synthesis that revises and updates the classification and updates in accord with recent developments in European vegetation science [@RN4698] and habitat classification [@RN5061]. In this article we have performed such a synthesis with two main objectives: (1) provide an updated classification of anthropogenic plant communities in the Iberian Atlantic ecoregion, and (2) describe the main patterns of vegetation diversity in terms of floristic origin (natives, archaeophytes, neophytes); life forms; functional traits (plant height and flowering time); and ecological indicator values (temperature, moisture, light, nutrients, soil reaction, disturbance frequency and disturbance severity).

@@

We performed all data management and analysis with R version 4.3.1 [@RN5387], using the R package *tidyverse* [@RN4662] for data processing and visualization. We homogenized all taxon names in the analysis and the article using Euro+Med [@RN5044] or Plants of the World Online for taxa not in Euro+Med [@RN5376].  For the nomenclature of sintaxa we follow Mucina *et al.* [-@RN4698] for alliances and higher ranks, and Rivas-Martínez *et al.* [-@RN4697] for associations. The original datasets, as well as R code for analysis and creation of the manuscript can be accessed at the GitHub repository <https://github.com/efernandezpascual/manmade>.

@@

We studied the anthropogenic plant communities of the Iberian Atlantic ecoregion [@RN5372], i.e. the territories with a temperate climate in the north-western Iberian Peninsula. Our study ecoregion broadly corresponds with the Cantabrian Mixed Forests ecoregion *sensu* Olson *et al.* [-@RN4667], with the Iberian part of the European Atlantic province *sensu* Rivas-Martínez *et al.* [-@RN4699] and with the Iberian section of the Atlantic biogeographical region of the European Environmental Agency (<https://www.eea.europa.eu/data-and-maps/figures/biogeographical-regions-in-europe-2>).

@@

To circumscribe our study vegetation, we followed the definition of anthropogenic vegetation in the recent classification of the vegetation of Europe by Mucina *et al.* [-@RN4698]. In our study area, this potentially includes the vegetation classes *Polygono-Poetea annuae*, *Papaveretea rhoeadis*, *Digitario sanguinalis-Eragrostietea minoris*, *Chenopodietea*, *Sisymbrietea*, *Bidentetea*, *Artemisietea vulgaris* and *Epilobietea angustifolii*. To these, and for a complete synthesis, we added to our study the class *Cymbalario-Parietarietea diffusae* included by Mucina *et al.* [-@RN4698] in the vegetation of rock crevices and screes, since this class encompasses the vegetation of human-made walls.

## Checklist of anthropogenic sintaxa

To assist in our classification, we prepared a checklist of anthropogenic sintaxa that could potentially be present in our study ecoregion (**Supplementary 1**), based on regional sintaxonomical checklists [@RN5363; @RN5008; @RN5365] and recent revisions at the European level [@RN4698; @RN5371]. The checklist inclused 38 anthropogenic vegetation alliances that could be present in the Iberian Atlantic ecoregion (**Supplementary 2**). A goal of our study was to validate the presence of these alliances in our ecoregion, and to assign to them the existing vegetation data.

@@

We retrieved anthropogenic vegetation plots from SIVIM, the Iberian and Macaronesian Vegetation Information System [@RN5374]. To identify plots as anthropogenic, we selected those plots that had been classified by the original authors of the plot as belonging to any sintaxa in our sintaxonomical checklist. However, since not all plots in SIVIM have an assigned sintaxon, we also retrieved additional plots using the expert system created by Chytrý *et al.* [-@RN5061] to classify vegetation plots into EUNIS pan-European habitat types. This expert system uses the floristic composition of plots to assign them to habitat types depending on the presence and abundance of characteristic species. We retrieved plots that had been assigned to any habitat related to the vegetation classes we had previously defined: all habitats in the level 1 code V (vegetated man-made habitats), plus level 3 codes R55 (lowland moist or wet tall-herb and fern fringe) and R57 (herbaceous forest clearing vegetation). It must be noted that habitats R55 and R57 include communities that are classified as *grasslands and lands dominated by forbs, mosses or lichens * by EUNIS but as *anthropogenic vegetation* by by Mucina *et al.* [-@RN4698]. The extraction from SIVIM produced an initial pool of 3,160 vegetation plots, to which we added 89 vegetation plots of urban or peri-urban plant communities sampled by us or extracted from local literature not in SIVIM [@RN5377; @RN5378].

@@

We performed an exploratory analysis of these 3,249 plots using Two-Way Indicator Species Analysis (TWINSPAN) [@RN5379]. We used the R package *twinspanR* [@RN5380] to do the TWINSPANs in this article, using 3 pseudospecies cut levels (0, 15, 25) with a minimum group size of 10 plots and Sørensen's average dissimilarity. During this exploratory analysis we identified 741 outlier plots which either did not belong to the target anthropogenic vegetation or to the geographical area of the ecoregion. The majority of these plots corresponded to coastal plant communities and had been misclassified by the expert system. We removed these, leaving a dataset of 2,508 vegetation plots for data analysis: 2,419 plots from SIVIM, originally recorded in 89 publications (**Supplementary 3**); plus 89 plots added by us.

@@

To create a training dataset for semi-supervised classification (see below), we started by keeping only those plots (n = 2,201) that had been assigned by the original authors of the plot as belonging to any of the 38 alliances in our sintaxonomical checklist (i.e., we removed the plots without sintaxa, that had been retrieved solely by the expert system). We further cleaned this subset by subjecting it to a preliminary numerical classification using TWINSPAN, in which we attempted to match the plots into the 38 alliances our sintaxonomical checklist. This step allowed us to identify those plots in which there was an agreement between (1) our TWINSPAN-based classification and (2) the classification based on the original-author sintaxa; these plots (n = 1,725) became our training dataset for semi-supervised classification. Furthermore, this step allowed us to validate which alliances from our sintaxonomical checklist were present in the study ecoregion (n = 28) and which were absent (n = 10).

@@

Next, we conducted a semi-supervised classification of the whole dataset (n = 2,508 plots) into the 28 validated anthropogenic alliances. Semi-supervised classification uses a training subset of *a priori* classified vegetation plots to classify a secondary subset of unclassified plots [@RN5383]. Since our goal was to refine the classification of the whole dataset, we allowed plots from the training subset to be re-assigned to other alliances during the classification. In addition, semi-supervised classification can create new groups to place data points that do not match the already existing *a priori* groups, but attempts to do so resulted in new groups with no ecological significance, and thus we kept the  28 alliances as final vegetation groups. Furthermore, we applied a noise clustering fuzzy algorithm, which allows the plots to be classified either into the *a priori* alliances or into a *noise* group which includes outliers and transitional plots [@RN5382]. We set the fuzziness coefficient to a low value (m = 1) to accommodate a high number of transitional plots; and we set the distance to the noise class to d = 1. We performed this analysis with the R package *vegclust* [@RN5383]. The semi-supervised classification with noise clustering resulted in the final classification of 2,081 plots into 28 alliances and 427 plots left out in the noise group. We used pairwise PERMANOVA (with 100,000 iterations, Euclidean distances and Holm's p-value correction) fitted with the R package *RVAideMemoire* [@RN5381] to test the significance of the final vegetation alliances; along with Principal Component Analysis (PCA) as implemented in the R package *FactoMineR* [@RN3166] to visualize the relationships between the alliances.

## Characteristic species and EUNIS habitat regionalization

We calculated sets of characteristic species [@RN5061] for each of the final 28 vegetation alliances. We defined dominant species as species with more than 25% cover in at least 5% of the vegetation plots of the group; constant species as species with a frequency higher than 50% in the group; and diagnostic species as species with an *IndVal* value with a p-value lower than 0.05, as calculated with the R package *labdsv* [@RN2311] using 1.000.000 iterations. Additionally, we assigned to each alliance a regionalized level 4 EUNIS habitat code, using pre-existing codes when they existed (<https://eunis.eea.europa.eu/habitats-code-browser-revised.jsp>), and creating new codes when not.

## Species origin as native, archaeophyte and neophyte

We classified the species as native, archaophytes and neophytes using the information in *Flora iberica* [-@RN3243] and catalogues of archaeophytes for Britain [@RN5490] and the Czech Republic [@RN5495]. We must stress that identifying archaeophytes is southern Europe is highly problematic and our classification must be taken as an indication of putative archaeophyte character for the purposes of vegetation description, rather than a definitive classification of the species.

@@

From *FloraVeg.EU* [-@RN5385] we extracted the species’ life forms and kept *therophyte* and *geophyte* for further analysis since these two categories had the largest contribution to variance as per an exploratory Principal Component Analysis (PCA) performed using the R package *FactoMineR* [@RN3166].

## Species and vegetation height and flowering phenology

From *Flora iberica* [-@RN3243] we extracted the maximum height, median month of flowering and length of the flowering period of each species in the dataset. We used these to calculate plot-level community-weighted means (weighting by species cover) for each trait, to characterize the vegetation height and flowering phenology of the anthropogenic plant communities.

@@

We collected, for each species in the dataset, its ecological indicator values of temperature, moisture, light, nutrients and soil reaction [@RN5384]; and its disturbance frequency and severity indicator values [@RN5101]. To those species lacking an indicator value in the accessed references, we assigned a value by reciprocal averaging, i.e. by (1) calculating weighted average values of the species with values for each plot (weighting by species cover), and (2) assigning to the missing species the weighted average values of the plots where they were present (weighting by species cover). Then, for each vegetation plot, we calculated the plot-level mean of each indicator value.

@@

The semi-supervised classification resulted in the classification of 2,081 vegetation plots into 28 anthropogenic alliances of the Iberian Atlantic ecoregion (**Table 1**, **Figure 1**). Most of the alliances had a wide distribution in the ecoregion according to our dataset, although five of them (*Caucalidion lappulae*, *Linario polygalifoliae-Vulpion alopecuri*, *Convolvulo arvensis-Agropyrion repentis*, *Senecionion fluviatilis*, *Paspalo-Agrostion semiverticillati*) had only isolated occurrences in a few points towards the southern limit of the Iberian Atlantic territories. The regional literature included in our dataset recognized 72 associations (**Supplementary 4**) within these alliances.

Focusing on the classified plots, the dataset included 1,167 taxa or taxa aggregates. Of these, 101 were neophytes (i.e. 9% of the species pool). Regarding life forms, the species pool was dominated by hemicryptophytes (43%) and therophytes (41%). The 10 most frequent species were *Ochlopoa annua* (621 occurrences), *Urtica dioica* (592), *Sonchus oleraceus* (540), *Stellaria media* (516), *Capsella bursa-pastoris* (390), *Polygonum aviculare* (375), *Dactylis glomerata* (363), *Senecio vulgaris* (332) and *Anisantha sterilis* (301). Considering only those plots with the most frequent plot size (20 m^2^, n = 256 plots), the average species richness per plot was 16 (minimum = 3, maximum = 37).

DCA ordination of the floristic composition of the classified plots (**Fig. 2**) suggested a high consistency of the vegetation classes *Cymbalario-Parietarietea diffusae*, *Polygono-Poetea annuae*, *Papaveretea rhoeadis*, *Digitario sanguinalis-Eragrostietea minoris*, *Sisymbrietea* and *Bidentetea*. The alliances belonging to the two perennial ruderal vegetation classes (*Artemisietea vulgaris* and *Epilobietea angustifolii*) were mixed together. The winter-annual class *Chenopodietea* also showed high heterogeneity, especially because of the isolation of the altered dune alliance *Linario polygalifoliae-Vulpion alopecuri*.

## Community-level traits and ecological preferences

PCA ordination of the community-weighted means of the species traits (**Fig. 3A**) indicated that the first two axes explained 65% of the variability. The major contributors to axis 1 (41% variance explained) were community height, proportion of geophytes, mean month of flowering and proportion of therophytes; it separated taller communities with more geophytes and later flowering (classes *Artemisietea vulgaris*, *Epilobietea angustifolii* and *Bidentetea*) from shorter communities with more therophytes and earlier flowering (classes *Polygono-Poetea annuae*, *Papaveretea rhoeadis* and *Sisymbrietea*). The major contributors to axis 2 (24% variance explained) were proportion of neophytes and width of the flowering season; it separated communities with more neophytes and longer flowering seasons (classes *Bidentetea*, *Digitario sanguinalis-Eragrostietea minoris* and *Polygono-Poetea annuae*) from communities with less neophytes and shorter flowering seasons (classes *Artemisietea vulgaris*, *Epilobietea angustifolii* and *Chenopodietea*).

>>>>

---
title: "Classification and characterization of anthropogenic plant communities at the ecoregion level"
output: 
  bookdown::word_document2:
    reference_docx: template.docx
    number_sections: false
bibliography: refs.bib
csl: journal-of-vegetation-science.csl
link-citations: yes
always_allow_html: true
---

@@

# Abstract (300 words max.)

* *Questions* How can we classify the diversity of anthropogenic plant communities at the ecoregion level? How are these communities characterized by species origins (native plants, archaeophytes, and neophytes), species traits (lifeforms, plant height, and flowering phenology) and ecological preferences (temperature, moisture, light, nutrients, soil reaction, disturbance frequency, and disturbance severity)?

* *Location* Iberian Atlantic ecoregion (a.k.a. Cantabrian Mixed Forests); Portugal, Spain, and France; south-western Europe.

* *Methods* We compiled a vegetation database of 2,508 synanthropic plots and revised the regional checklist of anthropogenic vegetation alliances. We used modified TWINSPAN and semi-supervised classification to classify the plots into the revised alliances. We determined the proportion of natives, archaeophytes, and neophytes. We also described the alliances in terms of species traits and ecological requirements.

* *Results* We classified 2,081 vegetation plots into 28 anthropogenic alliances representing 9 vegetation classes (*Cymbalario-Parietarietea diffusae*, *Polygono-Poetea annuae*, *Papaveretea rhoeadis*, *Digitario sanguinalis-Eragrostietea minoris*, *Chenopodietea*, *Sisymbrietea*, *Bidentetea*, *Artemisietea vulgaris* and *Epilobietea angustifolii*). The plots included 1,162 plant taxa: 78% natives, 15% archaeophytes, and 7% neophytes. Vegetation groups were organized along a principal axis of variation related to abiotic stress (dry-sunny to moist-shady habitats), and a second axis related to disturbance (low to high disturbance frequency and severity).

* *Conclusions* Ecoregion-level synthesis detected a discrepancy between the number of anthropogenic vegetation units described in the literature (38 alliances) and the number supported by numerical classification (28). The diversity of anthropogenic vegetation can be organized into three groups: trampled, weeds, and ruderals. In the Iberian Atlantic ecoregion, anthropogenic habitats host one third of the ecoregion plant species pool and one fifth of the Iberian flora. Mesic perennial ruderal vegetation is especially rich in native species and can be a biodiversity asset in urban ecosystems. Our ecoregion-level framework can improve the assessment and management of anthropogenic plant communities within biogeographically meaningful regions.

# Keywords (10 max)

Synanthropic vegetation, manmade habitats, human-made habitats, urban biodiversity, neophytes, archaeophytes, alien plants, habitat classification, semi-supervised classification, Iberian Atlantic ecoregion

@@

Anthropogenic vegetation consists of plant communities assembled as a direct consequence of human activities. In general, this vegetation is composed of so-called weeds of arable fields and ruderal plants of human settlements and nearby habitats [@RN5441]. Anthropogenic communities make up an important part of the supply of biodiversity in urban and peri-urban areas [@RN5445]. In densely populated landscapes, the plant communities that colonize urban vacant lots [@RN2300] are a useful natural resource with high ecological and societal potential [@RN5446]. Anthropogenic vegetation can also mitigate regional-level extreme climatic events [@RN5440]. Furthermore, many synanthropic plants can accumulate heavy metals in their biomass [@RN5433], making them valuable candidates for nature-based solutions such as bioremediation and restoration of degraded post-industrial landscapes [@RN5447].

In temperate Europe, anthropogenic vegetation is a melting pot of floras from different origins. Starting with a pool of native species favored by humans, and constantly enriched by human-assisted arrival of alien (i.e. non-native) species [@RN5504]. The introduction of archaeophytes (i.e. alien species arriving before 1.500 CE) likely started in the early Neolithic and peaked in the Bronze Age, with successive waves during periods of accelerated human colonization such as the early Middle Ages [@RN5504]. In Europe, synanthropic archaeophytes include many Mediterranean taxa originating from the segetal flora of the Near East [@RN5494]. Some of these archaeophytes are 'obligate weeds' that do not occur in non-anthropogenic habitats and are assumed to have evolved together with human agriculture [@RN5494]. Since many archaeophytes started their expansion into temperate areas during the Mid-Holocene warm period, it is difficult to disentangle to what extent this expansion was favored by human activities and/or ongoing climatic changes [@RN5432; @RN5443]. The synanthropic flora of Europe has been further enriched by neophytes, alien taxa that arrived after the expansion of intercontinental trade in the Modern Age (after 1.500 CE) [@RN5439]. Synanthropic neophytes therefore include many American, African, and Asian taxa that arrived in Europe because of the global trade networks established by western European colonialism [@RN5451]. In the last century, the European anthropogenic vegetation has shown a decrease in species richness and diversity [@RN5423], as rare synanthropic species have become rarer, and neophytes have increased their abundance at the expense of natives and archaeophytes [@RN5441].

Despite the potential of anthropogenic vegetation as a biodiversity asset in human-made landscapes, and its ecological interest as an assemblage of species pools from different biogeographical origins, human-made vegetation is largely absent from ecosystem management. A comprehensive management will require a solid definition, classification, and description of anthropogenic plant communities. However, the classification of anthropogenic vegetation is subject to particularities making it a special case compared to the classification of non-anthropogenic communities. In human-made habitats, the traditional understanding of community assembly based on sequential filtering (dispersal, abiotic, and biotic) must incorporate an additional filtering based on human preferences and actions [@RN5505]. Human landscapes are a mosaic of land uses [@RN5527], having specific effects on plant community assembly, and driven by sociological rather than traditional ecological factors [@RN2300]. In addition to these potentially confounding factors, human-assisted dispersal means that anthropogenic floras are relatively homogeneous across large biogeographical scales. However, traditional phytosociology has tended to classify them with a level of detail similar to the one dedicated to non-anthropogenic vegetation, leading to an inflation of anthropogenic syntaxa [@RN5528]. Another issue is defining anthropogenic vegetation, especially in a continent such as Europe where human impact on the landscape has been widespread since ancient times. The two major vegetation/habitat classifications coexisting in Europe today, phytosociological classification [@RN4698] and the EUNIS habitat classification [@RN5061], have non-overlapping definitions of which plant communities fall within the ‘anthropogenic’ or ‘manmade’ categories. These issues are further muddled by the fact that most classification studies of anthropogenic vegetation have focused either on a restricted political unit [@RN5424; @RN5426] or in a subset of the vegetation, such as segetal [@RN5494], ruderal [@RN5422] or trampled [@RN5434] vegetation. There is a need for a numerical classification and synthesis of the full spectrum of anthropogenic vegetation, and an appropriate scale to conduct such study is the ecological region (ecoregion, hereafter), since these biogeographical units are assumed to encompass areas with a similar biogeographic history and recurrent local ecosystems [@RN5491].

The Iberian Atlantic ecoregion (a.k.a. Cantabrian Mixed Forests ecoregion; NW Iberian Peninsula) has a long history of human habitation dating back to the Cantabrian Upper Paleolithic [@RN5450]. This, together with its transitional position at the border between the temperate and Mediterranean climatic zones of Europe [@RN4700], suggests a long regional history for archaeophyte-rich anthropogenic plant communities, supporting the current diverse anthropogenic vegetation with both temperate and Mediterranean floristic elements [@RN5008]. Moreover, historical trade links with America and Asia, together with the warm and humid temperate climate, have made the region a hotspot for biological invasions [@RN5492; @RN5493]. Recent post-industrial land-use changes have left large areas with abandoned industrial sites (i.e. brownfields) whose management and restoration requires an understanding of the anthropogenic communities able to colonize them [@RN5448; @RN5449]. Although there is a long tradition of studying weed and ruderal communities in the ecoregion [@RN5364; @RN5361; @RN5359; @RN5360] there is a current need for an ecoregion-level synthesis that revises and updates the classification in accord with recent developments in European vegetation science [@RN4698] and habitat classification [@RN5061]. In this article, we have performed such a synthesis with the following two objectives: (1) to provide an updated classification of anthropogenic plant communities in the Iberian Atlantic ecoregion; and (2) to characterize the diversity of anthropogenic vegetation in terms of species pools (natives, archaeophytes, neophytes); species traits (life form, plant height, flowering phenology); and ecological requirements (temperature, moisture, light, nutrients, soil reaction, disturbance frequency, and disturbance severity).

@@

## Data curation and taxonomic criteria

We performed all data management and analysis with R version 4.3.1 [@RN5387], using the R package *tidyverse* [@RN4662] for data processing and visualization. We homogenized all taxon names using Euro+Med [@RN5044] or Plants of the World Online [@RN5376] for taxa not in Euro+Med. For the nomenclature of syntaxa, we followed Mucina *et al.* [-@RN4698] for alliances and higher ranks, and Rivas-Martínez *et al.* [-@RN4697] for associations. The original datasets, as well as R code for analysis and creation of the manuscript, are stored in a GitHub repository available at Zenodo (see 'Data availability statement').

@@

We studied the anthropogenic plant communities of the Iberian Atlantic ecoregion [@RN5372], i.e. the territories with a temperate climate in the north-western part of the Iberian Peninsula (NW Portugal, N Spain, SW France). Our study ecoregion broadly corresponds to the Cantabrian Mixed Forests ecoregion *sensu* Olson *et al.* [-@RN4667], to the Iberian part of the European Atlantic province *sensu* Rivas-Martínez *et al.* [-@RN4699] and to the Iberian section of the Atlantic biogeographical region of the European Environmental Agency (<https://www.eea.europa.eu/data-and-maps/figures/biogeographical-regions-in-europe-2>).

@@

To circumscribe our study vegetation, we followed the definition of anthropogenic vegetation in the revised classification of the vegetation of Europe by Mucina *et al.* [-@RN4698]. In our study area, this potentially includes the vegetation classes *Polygono-Poetea annuae*, *Papaveretea rhoeadis*, *Digitario sanguinalis-Eragrostietea minoris*, *Chenopodietea*, *Sisymbrietea*, *Bidentetea*, *Artemisietea vulgaris*, and *Epilobietea angustifolii*. For the sake of completeness, we also considered the class *Cymbalario-Parietarietea diffusae* included by Mucina *et al.* [-@RN4698] in the vegetation of rock crevices and screes, since this class encompasses the vegetation of human-made walls.

## Checklist of anthropogenic syntaxa

To assist in our classification, we prepared a checklist of anthropogenic syntaxa that could be present in our study ecoregion (**Appendix S1**), based on regional syntaxonomical checklists [@RN5363; @RN5008; @RN5365] and recent revisions at the European level [@RN4698; @RN5371]. This syntaxonomical checklist included 38 anthropogenic vegetation alliances that could be present in the Iberian Atlantic ecoregion according to the literature.

@@

We retrieved Iberian Atlantic anthropogenic vegetation plots from the Iberian and Macaronesian Vegetation Information System (SIVIM) [@RN5374]. To identify plots as anthropogenic, we selected those plots that had been assigned by the original authors of the plot into any anthropogenic syntaxa in our syntaxonomical checklist. However, since not all plots in SIVIM have an assigned syntaxon, we also retrieved additional plots using the expert system created by Chytrý *et al.* [-@RN5061] to classify vegetation plots into EUNIS pan-European habitat types. We retrieved plots that had been assigned to any habitat related to the anthropogenic vegetation classes we had previously defined: all habitats in the level 1 code *V Vegetated man-made habitats*, plus level 3 codes *R55 Lowland moist or wet tall-herb and fern fringe* and *R57 Herbaceous forest clearing vegetation*. It must be noted that habitats *R55* and *R57* include communities that are classified as *R Grasslands and lands dominated by forbs, mosses or lichens* by EUNIS but as *Anthropogenic vegetation* by Mucina *et al.* [-@RN4698]. The extraction of data from SIVIM produced an initial pool of 3,160 vegetation plots, to which we added 89 vegetation plots of urban or peri-urban plant communities sampled by us or extracted from local literature not included in SIVIM [@RN5377; @RN5378].

@@

We performed an exploratory data analysis of these 3,249 plots using modified Two-Way Indicator Species Analysis (TWINSPAN) [@RN5379] with the R package *twinspanR* [@RN5380], 3 pseudospecies cut levels (0, 15, 25), a minimum group size of 10 plots, and Sørensen's average dissimilarity. During this analysis we identified 741 outlier plots which did not belong to either the target anthropogenic vegetation or to the geographical area of the ecoregion. The majority of these plots corresponded to coastal plant communities and had been misclassified by the expert system. We removed these plots, leaving a dataset of 2,508 vegetation plots for data analysis: 2,419 plots from SIVIM, originally recorded in 89 publications (**Appendix S2**); plus 89 plots added by us.

@@

To create a training dataset for the semi-supervised classification (see below), we started by keeping only those plots (n = 2,201) that had been assigned by the original authors of the plot to any of the 38 anthropogenic alliances defined in our syntaxonomical checklist (i.e., we removed the plots without syntaxa, that had been retrieved solely by the expert system). We further cleaned this subset using a preliminary TWINSPAN classification, in which we attempted to match the plots into the 38 alliances. This step allowed us to identify those plots in which there was an agreement between (1) our TWINSPAN-based classification and (2) the classification based on the original author's syntaxon. These plots (n = 1,725) became our training dataset for semi-supervised classification. Furthermore, this step allowed us to validate which alliances from our syntaxonomical checklist were present in the study ecoregion (n = 28) and which were absent (n = 10).

@@

Next, we conducted a semi-supervised classification of the whole dataset (n = 2,508 plots) into the 28 validated anthropogenic alliances. Semi-supervised classification uses a training subset of *a priori* classified vegetation plots to classify a secondary subset of unclassified plots [@RN5383]. Since our goal was to refine the classification of the whole dataset, we allowed plots from the training subset to be re-assigned to other alliances during the classification. In addition, semi-supervised classification can create new groups to place data points that do not match the already existing *a priori* groups, but attempts to do so resulted in new groups with no ecological significance, and thus we kept the 28 alliances as final anthropogenic vegetation groups. Furthermore, we applied a noise clustering fuzzy algorithm, which allows plots to be classified either into the *a priori* alliances or into a *noise* group which includes outliers and transitional plots [@RN5382]. We set the fuzziness coefficient to a low value (m = 1) to accommodate a high number of transitional plots; and we set the distance to the noise class to d = 1. We performed this analysis with the R package *vegclust* [@RN5383]. The semi-supervised classification with noise clustering resulted in the final classification of 2,081 plots into 28 alliances and 427 plots left out in the noise group. We used pairwise PERMANOVA (with 100,000 iterations, Euclidean distances, and Holm's p-value correction) fitted with the R package *RVAideMemoire* [@RN5381] to test the significance of the final vegetation alliances; along with Principal Component Analysis (PCA) as implemented in the R package *FactoMineR* [@RN3166] to visualize the relationships between the classes and alliances.

## Characteristic species and EUNIS habitat correspondence

To facilitate the use of the classification by local managers and environmental consultants, we calculated sets of characteristic species (**Appendix S3**) [@RN5061] for each one of the final 28 vegetation alliances. We defined dominant species as species with more than 25% cover in at least 5% of the vegetation plots of the group; constant species as species with a frequency higher than 50% in the group; and diagnostic species as species whose *IndVal* had a p-value lower than 0.05, as calculated with the R package *labdsv* [@RN2311] using 1.000.000 iterations. Additionally, we assigned to each alliance its corresponding level-3 EUNIS habitat code or codes (<https://eunis.eea.europa.eu/habitats-code-browser-revised.jsp>).

## Species origin as native, archaeophyte, or neophyte

We classified the species as native, archaeophytes, or neophytes in the Iberian Atlantic ecoregion using the information in *Flora iberica* [-@RN3243] and catalogues of archaeophytes for Britain [@RN5490] and the Czech Republic [@RN5495]. We must stress that identifying archaeophytes in southern Europe is highly problematic [@RN5444] and our classification must be taken as an indication of putative archaeophyte character for the purposes of vegetation description and comparison with other European regions, rather than a definitive classification of the species; as stated by Preston *et al.* [@RN5490], the classification of a species as an archaeophyte should be interpreted as a hypothesis to be tested by further studies.

@@

Using *FloraVeg.EU* [-@RN5385] we extracted the species’ life forms and kept *therophyte* and *geophyte* for further analysis since these two categories had the largest contribution to variance among alliances, as per an exploratory Principal Component Analysis (PCA) performed using the R package *FactoMineR* [@RN3166].

## Species height and flowering phenology

From *Flora iberica* [-@RN3243] we extracted the maximum height, median month of flowering, and length of the flowering period of each species in the dataset. We used these variables to calculate plot-level community-weighted means (weighting by species cover) for each trait, to characterize the vegetation height and flowering phenology of the anthropogenic plant communities.

@@

For each species in the dataset, we collected its ecological indicator values of temperature, moisture, light, nutrients, and soil reaction [@RN5488]; and its disturbance frequency and severity indicator values [@RN5101]. To those species lacking an indicator value in the accessed references, we assigned a value by reciprocal averaging, i.e. by (1) calculating weighted average values of the species with values for each plot (weighting by species cover), and (2) assigning to the missing species the weighted average values of the plots where they were present (weighting by species cover). Then, for each vegetation plot, we calculated the plot-level mean of each indicator value.

@@

The semi-supervised classification resulted in the classification of 2,081 vegetation plots into 28 anthropogenic alliances (**Fig. 1**, **Appendix S4**) representing 9 vegetation classes. Most of the alliances had a wide distribution in the ecoregion according to our dataset, but five of them (*Caucalidion lappulae*, *Linario polygalifoliae-Vulpion alopecuri*, *Convolvulo arvensis-Agropyrion repentis*, *Senecionion fluviatilis* and *Paspalo-Agrostion semiverticillati*) had isolated occurrences towards the southern limit of the Iberian Atlantic territories. The regional literature included in our dataset recognized 72 associations (**Appendix S5**) within these alliances, but exploratory analysis (not shown) indicated that the separation of these associations within the alliances had a generally weak support based on numerical classification methods.

The classified plots included 1,162 taxa or taxa aggregates. The 10 most frequent species were *Ochlopoa annua* (621 occurrences), *Urtica dioica* (592), *Sonchus oleraceus* (540), *Stellaria media* (516), *Capsella bursa-pastoris* (390), *Polygonum aviculare* (375), *Dactylis glomerata* (363), *Senecio vulgaris* (332), and *Anisantha sterilis* (301). Considering only those plots with the most frequent plot size range (10-30 m^2^, n = 867 plots), the average species richness per plot was 16 (minimum = 3, maximum = 48). The class with the richest species pool was *Epilobietea angustifolii* (n = 615) and the poorest was *Cymbalario-Parietarietea diffusae* (n = 171).

PCA ordination of the floristic composition (**Fig. 2A**) clearly differentiated three vegetation groups: (1) trampled vegetation of class *Polygono-Poetea annuae*; (2) crop weeds of classes *Papaveretea rhoeadis* and *Digitario sanguinalis-Eragrostietea minoris*; and (3) perennial ruderal vegetation of classes *Artemisietea vulgaris* and *Epilobietea angustifolii*. The central position of the floristic space, between these three major groups, was occupied by the annual ruderal vegetation of classes *Chenopodietea*, *Sisymbrietea*, and *Bidentetea*; as well as by the wall vegetation of class *Cymbalario-Parietarietea diffusae*. Separate PCAs within the major vegetation groups indicated a coherent separation between most of the alliances, but with some cases of relatively high overlap: for example, between (i) *Galio valantiae-Parietarion judaicae* and *Cymbalario-Asplenion*; or (ii) *Geo urbani-Alliarion officinalis* and *Aegopodion podagrariae*.

## Proportion of natives, archaeophytes and neophytes

The total synanthropic species pool (n = 1,162) was composed of 78% natives (n = 908), 15% putative archaeophytes (n = 178), and 7% neophytes (n = 76). The most frequent putative archaeophytes were *Capsella bursa-pastoris* (390 occurrences), *Senecio vulgaris* (332), *Anisantha sterilis* (301), *Anthemis arvensis* (280), *Hordeum murinum* (274), *Raphanus raphanistrum* (233), *Vicia sativa* (228), *Malva sylvestris* (197), *Sisymbrium officinale* (166), and *Echinochloa crus-galli* (165); while the most frequent neophytes were *Veronica persica* (212), *Erigeron canadensis* (138), *Amaranthus hybridus* (73), *Oxalis corniculata* (67), *Lepidium didymum* (57), *Oxalis latifolia* (51), *Paspalum distichum* (48), *Bidens frondosus* (35), *Cyperus eragrostis* (34), and *Symphyotrichum squamatum* (33).
